---
title: "6372 Project 1 - Life Expectancy"
author: "By Kristi Herman, Bo Yun, Eric Romero, Neil Benson"
date: "09/20/2020"
output:
  html_document:
    df_print: paged
---
  
# Introduction
Introduction paragraphs here


## Importing the necessary libraries
```{r,warning=FALSE,message=FALSE}

library(ggplot2)
library(ggthemes)
library(gridExtra)
library(tidyverse)
library(plyr)
library(dplyr) 
library(naniar) #keep
library(zoo)


# create not in operator
`%notin%` <- Negate(`%in%`)

```
  
  
## Importing and Cleaning the Data
```{r, warning=FALSE, message=FALSE}
########### Loading the data
life_ex_DF <- read.csv("https://raw.githubusercontent.com/kristxh/Applied_Stats_Project1/master/datasets_12603_17232_Life_Expectancy_Data.csv")

# cleaning up column names
colnames(life_ex_DF) <- tolower(colnames(life_ex_DF))
colnames(life_ex_DF) <- gsub("\\.\\.",".",colnames(life_ex_DF))
colnames(life_ex_DF) <- gsub("\\.","_",colnames(life_ex_DF))
names(life_ex_DF)[19] <- "thinness_10_19_years"

# converting status and year to factor columns
life_ex_DF$status <- as.factor(life_ex_DF$status)
```


## Checking for missing data
```{r, fig.height=7, fig.width=9}
# checking for missing values in the data
vis_miss(life_ex_DF) + xlab("Data Columns")
```
  
  
## Fix missing data
```{r, fig.height=7, fig.width=9}
# sorting the columns so that life_expectancy column is first
life_ex_DF <- life_ex_DF %>% select("life_expectancy", everything())

factorcols <- c("status","country","year")
allcols <- colnames(life_ex_DF)

# selecting only numerical columns for correlation
corvars <- allcols[allcols %notin% factorcols]


# impute with median missing values per country
life_ex_median_DF <- as.data.frame(life_ex_DF %>%
  group_by(country) %>%
  mutate_at(corvars, na.aggregate, FUN=median))


# dropping rows where the country has a single entry/record in the dataset and their records are mostly incomplete
life_ex_median_DF <- life_ex_median_DF[!is.na(life_ex_median_DF$life_expectancy), ]


# imputing the median by country still left a few nulls throughout. We will then impute the rest of the missing values as the mean of status (developing, developed, by year)
life_ex_cleaned_DF <- as.data.frame(life_ex_median_DF %>%
  group_by(status,year) %>%
  mutate_at(corvars, na.aggregate, FUN=median))

# checking for missing values in the data
vis_miss(life_ex_cleaned_DF) + xlab("Data Columns")
```


## Summary of cleaned up data
```{r}
summary(life_ex_cleaned_DF)
str(life_ex_cleaned_DF)

t(aggregate(life_expectancy~status,data=life_ex_cleaned_DF,summary))
t(aggregate(life_expectancy~status,data=life_ex_cleaned_DF,sd))
```
  
BMI has some very weird mins and maxes. Will explore further, but this data may later show correlations that aren't actually there because the min and max of BMI are not practical or possible.  


## BMI Boxplot by status (developing, developed)
```{r, fig.height=7, fig.width=9}
bmi_box <- geom_boxplot(
  mapping = aes(y=bmi),
  data = life_ex_cleaned_DF,
  colour = "black",
  fill = "#56B4E9",
  groupColors=c('#999999','#E69F00')
)

ggplot()+ bmi_box + xlab("BMI") + ggtitle("BMI")

```

## Histogram of life_expectancy
```{r, fig.height=7, fig.width=9}
theme_set(theme_igray())

le_hist <- geom_histogram(
  mapping = aes(life_expectancy),
  data = life_ex_cleaned_DF,
  bins = 20,
  colour = "gray",
  fill = "#0072B2"
)

ggplot()+ le_hist + xlab("Years") + ggtitle("Life Expectancy")
```


## Overall Life Expectancy
```{r, fig.height=7, fig.width=9}

le_by_yr <- ddply(life_ex_cleaned_DF, .(year), summarize,  avg_le=mean(life_expectancy))

le_line <- geom_line(
  mapping = aes(x = year, y = avg_le),
  data = le_by_yr,
  size = 1,
  colour = "dodgerblue4"
)

ggplot() + le_line + xlab("Year") + ylab("Life Expectancy") + ggtitle("Average Life Expectancy: 2000 - 2015") 

```

##################


Life Expectancy is linearly related to year, and year should be treated as continuous instead of as a factor in this model

DROP BMI for having unrealistic range of data

##################

## Developed/Developing Life Expectancy
```{r, fig.height=7, fig.width=9}
le_by_status <- ddply(life_ex_cleaned_DF, .(year,status), summarize,  avg_le=mean(life_expectancy))
line_colors <- c("dodgerblue4", "#D55E00")

le_line2 <- geom_line(
  mapping = aes(x = year, y = avg_le, group=status, color=status),
  data = le_by_status,
  size = 1
)

ggplot() + le_line2 +  xlab("Year") + ylab("Life Expectancy") + ggtitle("Average Life Expectancy: Developed and Developing Countries") + scale_colour_manual(values=line_colors)

```


## Boxplot by status (developing, developed)
```{r, fig.height=7, fig.width=9}
# Plot boxplot

dev_box <- geom_boxplot(
  mapping = aes(x=status, y=life_expectancy),
  data = life_ex_cleaned_DF,
  colour = "black",
  fill = "#56B4E9",
  groupColors=c('#999999','#E69F00')
)

ggplot()+ dev_box + xlab("Status of Country") + ylab("Life Expectancy") + ggtitle("Life Expectancy: Developed and Developing Countries ")

```

There appears to be enough of a difference between `life_expectancy` mean by developed and developing `status`. We will perform an anova and analyze signifance in the difference in means. 


## Boxplots by status by year
```{r, fig.height=7, fig.width=9}
# Plot boxplot

yr_box <- stat_boxplot(
  mapping = aes(x=year, y=life_expectancy, group=year),
  data = life_ex_cleaned_DF,
  colour = "black",
  fill = "#56B4E9"
  ) 
  

ggplot()+ yr_box + xlab("Year") + ylab("Life Expectancy") + scale_x_continuous(breaks=c(2000:2015)) + ggtitle("Life Expectancy by Year")  
```


## Checking the significance of the factorial variables using ANOVA
```{r}
# convert categorical variables to factors
names <- c("country", "year", "status")

life_ex_cleaned_factor_DF <- life_ex_cleaned_DF

# set the categorical columns to factor
for (name in names)
{
  life_ex_cleaned_factor_DF[,name]<-factor(life_ex_cleaned_DF[,name])
}

# convert factors to numeric for factor analysis
life_ex_cleaned_asnum_DF <- life_ex_cleaned_factor_DF[,names] %>% mutate_all(as.numeric)

# adding suffix ASNUM to numerical representation of factors columns
colnames(life_ex_cleaned_asnum_DF) <- paste(colnames(life_ex_cleaned_asnum_DF), "ASNUM", sep = "_")

# adding numerical factor columns to base Df in order to perform stepwise for variable selection
life_ex_cleaned_withnum_DF <- cbind(life_ex_cleaned_factor_DF, life_ex_cleaned_asnum_DF)

# dropping factor columns
life_ex_cleaned_withnum_DF <- life_ex_cleaned_withnum_DF[, sapply(life_ex_cleaned_withnum_DF, class) != "factor"]

#  columns for anova factor analysis
anovacols <- c(colnames(life_ex_cleaned_asnum_DF))

#  creating the formula to pass into aov function
anovafmla <- as.formula(paste("life_expectancy ~ ", paste(anovacols, collapse= "+")))

# passing my auto generated formula to analyze
anova_fit <- aov(anovafmla, data=life_ex_cleaned_withnum_DF)
summary(anova_fit)
```


### Testing the life_expectancy mean difference by status  
Equal means model significance here to test significance of mean difference of `life_expectancy` between status.  


H0: μdeveloped = μdeveloping  
Ha: μdeveloped ≠ μdeveloping  

Reject H0 that the means are equal  

This is statistically significant (p-value <2e-16), and we will include status as a factor to cconsider in the model.  

### Testing the life_expectancy mean difference by year  
Equal means model significance here to test significance of mean difference of `life_expectancy` between years.  


H0: μ2000 = μ2001 = μ2002 = μ2003 = μ2004 = μ2005 = μ2006 = μ2007 = μ2008 = μ2009 = μ2010 = μ2011 = μ2012 = μ2013 = μ2014 = μ2015  
Ha: μ2000 ≠ μ2001 ≠ μ2002 ≠ μ2003 ≠ μ2004 ≠ μ2005 ≠ μ2006 ≠ μ2007 ≠ μ2008 ≠ μ2009 ≠ μ2010 ≠ μ2011 ≠ μ2012 ≠ μ2013 ≠ μ2014 ≠ μ2015  

Reject H0 that the means are equal  

This is statistically significant (p-value <2e-16), and we will include year as a factor to cconsider in the model.  


### Both year and status are statistically significant as categorical predictors. 
Later will remove `country` from the data set for variable selection.  


## Scatterplot boxplot
```{r fig.width=10, fig.height=7}
le_scatter <- geom_point(
  mapping = aes(x=year, y=life_expectancy, color=status),
  data = life_ex_cleaned_DF
  ) 
  

ggplot()+ le_scatter + xlab("Year") + ylab("Life Expectancy") + scale_x_continuous(breaks=c(2000:2015)) + ggtitle("Life Expectancy: Developed and Developing Countries")  + scale_colour_manual(values=line_colors)
```


## Correlation matrix with pearson correlation coefficients
```{r, fig.height=9, fig.width=9}
library(reshape2)

corr <- round(cor(life_ex_cleaned_DF[,corvars]),1)
melted_cormat <- melt(corr)


# Get lower triangle of the correlation matrix
  get_lower_tri<-function(cormat){
    cormat[lower.tri(cormat)] <- NA
    return(cormat)
  }
  

# reorder the correlation matrix
reorder_cormat <- function(cormat){
# Use correlation between variables as distance
dd <- as.dist((1-cormat)/2)
hc <- hclust(dd)
cormat <-cormat[hc$order, hc$order]
}


# reorder the correlation matrix
corr <- reorder_cormat(corr)

# get lower half
lower_half <- get_lower_tri(corr)

# reshape the correlation matrix
melted_cormat <- melt(lower_half, na.rm = TRUE)

# plot it
ggcorrmap <- ggplot(melted_cormat, aes(Var2, Var1, fill = value))+
 geom_tile(color = "white")+
 scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
  midpoint = 0, limit = c(-1,1), space = "Lab", 
    name="Pearson\nCorrelation") +
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 45, vjust = 1, 
    size = 8, hjust = 1), axis.title.x = element_blank(), axis.title.y = element_blank())+
 coord_fixed()


ggcorrmap + geom_text(aes(Var2, Var1, label = value), color = "black", size = 2.5)
```
It appears there is a high degree of correlation between the following:  
- `thinnes_5_9_years` & `thinnes_10_19_years`  
- `percentage_expenditure` & `gdp`  
- `under_5_deaths` & `infant_deaths`  
- `schooling` & `income_composition_of_resources`  
- `population` has 0 correlation with the response variable and will be removed as a predictor and is incomplete throughout the data set  


## Further examining the issue of multicolinearity
```{r, fig.height=9, fig.width=9}
pairs(life_ex_cleaned_DF[,corvars], pch = 19,  cex = 0.5,
      col = line_colors[life_ex_cleaned_DF$status],
      upper.panel=NULL)

```
It appears that both `hiv_aids` and `percentage_expenditure` have nonlinear relationships with the response variable. We will further investigate.  

There also appears to be some interaction between `status` and `percentage_expenditure`, `status` and `schooling`, and `status` and `bmi`. Further exploration needed.


## Further exploration of hiv_ads and percentage_expenditure
```{r, fig.height=7, fig.width=9}
par(mfrow=c(1,2))

plot(life_ex_cleaned_DF$hiv_aids,life_ex_cleaned_DF$life_expectancy, xlab="hiv_aids",ylab="life_expectancy")


plot(life_ex_cleaned_DF$percentage_expenditure,life_ex_cleaned_DF$life_expectancy, xlab="percentage_expenditure",ylab="life_expectancy")
```


```{r, fig.height=7, fig.width=9}

life_expectancy <- life_ex_cleaned_DF$life_expectancy
hiv_aids <- life_ex_cleaned_DF$hiv_aids

hiv_aids.model <- lm(life_expectancy~hiv_aids)

# visually inspecting the reltationship between life_expectancy and hiv_aids before adding complexity
par(mfrow=c(1,3))
plot(hiv_aids,life_expectancy, xlab="hiv_aids",ylab="life_expectancy")
new<-data.frame(hiv_aids=seq(0,50,.1))
lines(seq(0,50,.1),predict(hiv_aids.model,newdata=new),col="red",lwd=4)
plot(hiv_aids.model$fitted.values,hiv_aids.model$residuals,xlab="Fitted Values",ylab="Residuals")
plot(life_ex_cleaned_DF$hiv_aids,hiv_aids.model$residuals,xlab="hiv_aids",ylab="Residuals")


# adding complexity to relationship between life_expectancy and hiv_aids
hiv_aids.model2 <- lm(life_expectancy~hiv_aids+I(log(hiv_aids)))


# visually inspecting the reltationship between life_expectancy and hiv_aids after transformations
par(mfrow=c(1,3))
plot(hiv_aids,life_expectancy, xlab="hiv_aids",ylab="life_expectancy")
new<-data.frame(hiv_aids=seq(0,50,.1))
lines(seq(0,50,.1),predict(hiv_aids.model2,newdata=new),col="red",lwd=4)
plot(hiv_aids.model2$fitted.values,hiv_aids.model2$residuals,xlab="Fitted Values",ylab="Residuals")
plot(life_ex_cleaned_DF$hiv_aids,hiv_aids.model2$residuals,xlab="hiv_aids",ylab="Residuals")

```


```{r, fig.height=7, fig.width=9}

percentage_expenditure <- life_ex_cleaned_DF$percentage_expenditure

perc_exp.model <- lm(life_expectancy~percentage_expenditure)

# visually inspecting the reltationship between life_expectancy and percentage_expenditure before adding complexity
par(mfrow=c(1,3))
plot(percentage_expenditure,life_expectancy, xlab="percentage_expenditure",ylab="life_expectancy")
new2<-data.frame(percentage_expenditure=seq(0,20000,100))
lines(seq(0,20000,100),predict(perc_exp.model,newdata=new2),col="red",lwd=4)
plot(perc_exp.model$fitted.values,perc_exp.model$residuals,xlab="Fitted Values",ylab="Residuals")
plot(life_ex_cleaned_DF$hiv_aids,perc_exp.model$residuals,xlab="hiv_aids",ylab="Residuals")


# adding complexity to relationship between life_expectancy and percentage_expenditure
perc_exp.model2 <- lm(life_expectancy~percentage_expenditure+I(sqrt(percentage_expenditure)))


# visually inspecting the reltationship between life_expectancy and percentage_expenditure after transformations
par(mfrow=c(1,3))
plot(percentage_expenditure,life_expectancy, xlab="percentage_expenditure",ylab="life_expectancy")
new2<-data.frame(percentage_expenditure=seq(0,20000,100))
lines(seq(0,20000,100),predict(perc_exp.model2,newdata=new2),col="red",lwd=4)
plot(perc_exp.model2$fitted.values,perc_exp.model2$residuals,xlab="Fitted Values",ylab="Residuals")
plot(life_ex_cleaned_DF$percentage_expenditure,perc_exp.model2$residuals,xlab="percentage_expenditure",ylab="Residuals")
```


## Checking colinearity with VIF and removing redundant columns
```{r}
library(car)

full.model <- lm(life_expectancy~. ,data=life_ex_cleaned_DF[,corvars])

vif(full.model, data = life_ex_cleaned_DF)

# removing columns and re-evaluating
colstodelete <- c("under_five_deaths", "thinness_5_9_years", "gdp", "population")

full.model1 <- lm(life_expectancy~. ,data=life_ex_cleaned_DF[,corvars] %>% select(-all_of(colstodelete)))

vif(full.model1, data = life_ex_cleaned_DF[,corvars] %>% select(-all_of(colstodelete)))

```
- `infant_deaths` and `under_five_deaths` have a VIF of 177.36 and 176.35 respectively, indicating a very high degree of multicolinearity between the two. We will remove `under_five_deaths` and re-evaluate.  
  
- `thinness_5_9_years` and `thinness_10_19_years` have a VIF of 8.82 and 8.74 respectively, indicating a very high degree of multicolinearity between the two. We will remove `thinness_5_9_years` and re-evaluate.  
  
- `percentage_expenditure` and `gdp` have a VIF of 4.92 and 5.32 respectively, indicating a very high degree of multicolinearity between the two. We will remove `gdp` and re-evaluate.  
  
- `schooling` and `income_composition_of_resources` are correlated to each other with a high Pearson Correlation Coefficient of .8, but have VIFs of 3.62 and 3.05 respectively. While these are highly correlated, their VIFs are relatively low. We will leave them in them in for variable selection.  
  

## Removing columns with high colinearity or low significance
"under_five_deaths", "thinness_5_9_years", "gdp", "population" and "country"  
```{r}
# adding "country" to the list of columns to delete after its poor performance in ANOVA analysis
colstodelete <- append(colstodelete, "country")  
Life <- life_ex_cleaned_factor_DF %>% select(-colstodelete) 
```


## Checking residuals and q-q plot of the full model
Including complexity for the terms `hiv_aids` and `percentage_expenditure`.  
```{r, fig.height=7, fig.width=9}
full.model <- lm(life_expectancy~. + I(hiv_aids^2) + I(sqrt(percentage_expenditure)) ,data=Life)
par(mfrow=c(2,2))
plot(full.model)
```


## Adding interaction and checking significance
Visual assessment of pair plots showed potential interactions between the following:  

- `status` >> `percentage_expenditure`  
- `status` >> `bmi`  
- `status` >> `schooling`
```{r, fig.height=7, fig.width=9}
full.modelwinter <- lm(life_expectancy~. + I(hiv_aids^2) + I(sqrt(percentage_expenditure)) + status:percentage_expenditure + status:bmi + status:schooling, data=Life)

# Re-assessing the residual plots and assumptions
par(mfrow=c(2,2))
plot(full.modelwinter)
summary(full.modelwinter)
```


## Splitting the data into test and training sets for cross validation
```{r}
# Cross Validation on Train vs Test 
# Splitting 2939 observations of the data into Training and Test set in the ratio of 70:30
set.seed(1234)
trainIndices = sample(seq(1:dim(Life)[1]),round(.7*dim(Life)[1]))
train = Life[trainIndices,]
test = Life[-trainIndices,]
dim(train)
dim(test)
```


## Creating a prediction function
```{r}
predict.regsubsets = function (object , newdata ,id ,...){
  form=as.formula (object$call [[2]])
  mat=model.matrix(form ,newdata )
  coefi=coef(object ,id=id)
  xvars=names(coefi)
  mat[,xvars]%*%coefi
}
```


## Variable selection - Forward 
```{r, fig.height=7, fig.width=9}
# Just the training set
library(leaps)

reg.fwd=regsubsets(life_expectancy~. + I(hiv_aids^2) + I(sqrt(percentage_expenditure)) + status:percentage_expenditure+status:bmi + status:schooling, data=train, method="forward", nvmax=24)

par(mfrow=c(1,3))

# BIC
bics<-summary(reg.fwd)$bic
plot(1:24,bics,type="l",ylab="BIC",xlab="# of predictors")
index<-which(bics==min(bics))
points(index,bics[index],col="red",pch=10)

# Adjusted R2
adjr2<-summary(reg.fwd)$adjr2
plot(1:24,adjr2,type="l",ylab="Adjusted R-squared",xlab="# of predictors")
index<-which(adjr2==max(adjr2))
points(index,adjr2[index],col="red",pch=10)

# RSS
rss<-summary(reg.fwd)$rss
plot(1:24,rss,type="l",ylab="train RSS",xlab="# of predictors")
index<-which(rss==min(rss))
points(index,rss[index],col="red",pch=10)
mtext("Variable selection - Forward", side = 3, line = -2, outer = TRUE)

# -> 15 predictors
```


## Variable selection - Backward
```{r, fig.height=7, fig.width=9}
# Just the training set
reg.bwd=regsubsets(life_expectancy~. + I(hiv_aids^2) + I(sqrt(percentage_expenditure)) + status:percentage_expenditure+status:bmi + status:schooling, data=train, method="backward", nvmax=24)

par(mfrow=c(1,3))

# BIC
bics2<-summary(reg.bwd)$bic
plot(1:24,bics2,type="l",ylab="BIC",xlab="# of predictors")
index2<-which(bics2==min(bics2))
points(index2,bics2[index2],col="red",pch=10)

# Adjusted R2
adjr2<-summary(reg.bwd)$adjr2
plot(1:24,adjr2,type="l",ylab="Adjusted R-squared",xlab="# of predictors")
index.r2<-which(adjr2==max(adjr2))
points(index.r2,adjr2[index.r2],col="red",pch=10)

# RSS
rss2<-summary(reg.bwd)$rss
plot(1:24,rss2,type="l",ylab="train RSS",xlab="# of predictors")
index.rss2<-which(rss2==min(rss2))
points(index.rss2,rss[index.rss2],col="red",pch=10)
mtext("Variable selection - Backward", side = 3, line = -2, outer = TRUE)

# -> 13 predictors
```


## Variable selection - Stepwise
```{r, fig.height=7, fig.width=9}
# Just the training set
reg.stp=regsubsets(life_expectancy~. + hiv_aids^2 + sqrt(percentage_expenditure) + status*percentage_expenditure + status*bmi + status*schooling, data=train, method="seqrep", nvmax=24)

par(mfrow=c(1,3))

# BIC
bics3<-summary(reg.stp)$bic
plot(1:24,bics,type="l",ylab="BIC",xlab="# of predictors")
index3<-which(bics3==min(bics3))
points(index3,bics[index3],col="red",pch=10)

# Adjusted R2
adjr3<-summary(reg.stp)$adjr
plot(1:24,adjr3,type="l",ylab="Adjusted R-squared",xlab="# of predictors")
index3<-which(adjr3==max(adjr3))
points(index3,adjr3[index3],col="red",pch=10)

# RSS
rss3<-summary(reg.stp)$rss
plot(1:24,rss3,type="l",ylab="train RSS",xlab="# of predictors")
index3<-which(rss3==min(rss3))
points(index3,rss[index3],col="red",pch=10)
mtext("Variable selection - Stepwise", side = 3, line = -2, outer = TRUE)

# -> 13 predictors
```


## Variable selection - LASSO
```{r, fig.height=7, fig.width=9}
library(glmnet)

x=model.matrix(life_expectancy~. + I(hiv_aids^2) + I(sqrt(percentage_expenditure)) + status:percentage_expenditure+status:bmi + status:schooling, train)[,-1]
y=log(train$life_expectancy)

xtest<-model.matrix(life_expectancy~. + I(hiv_aids^2) + I(sqrt(percentage_expenditure)) + status:percentage_expenditure+status:bmi + status:schooling, test)[,-1]
ytest<-log(test$life_expectancy)

grid=10^seq(10,-2, length =100)
lasso.mod=glmnet(x,y,alpha=1, lambda =grid)

cv.out=cv.glmnet(x,y,alpha=1) #alpha=1 performs LASSO
plot(cv.out)

bestlambda<-cv.out$lambda.min  #Optimal penalty parameter.  You can make this call visually.
lasso.pred=predict (lasso.mod ,s=bestlambda ,newx=xtest)

testMSE_LASSO<-mean((ytest-lasso.pred)^2)
testMSE_LASSO

coef(lasso.mod,s=bestlambda)

reg.lso <- lm(life_expectancy~. -infant_deaths -alcohol -hepatitis_b -measles -total_expenditure, data=Life)
```


## Cross validating the models on the test data
```{r}
testASEfwd<-c()
testASEbwd<-c()
testASEstp<-c()

for (i in 1:24){
  # predicting the forward model on test set
  predictionsfwd<-predict.regsubsets(object=reg.fwd,newdata=test,id=i)
  testASEfwd[i]<-mean((test$life_expectancy-predictionsfwd)^2)
  
  # predicting the backward model on test set
  predictionsbwd<-predict.regsubsets(object=reg.bwd,newdata=test,id=i)
  testASEbwd[i]<-mean((test$life_expectancy-predictionsbwd)^2)
  
  # predicting stepwise model on test set
  predictionsstp<-predict.regsubsets(object=reg.stp,newdata=test,id=i)
  testASEstp[i]<-mean((test$life_expectancy-predictionsstp)^2)
  
}

```


## Plots for something - ASE
```{r}
par(mfrow=c(1,1))
plot(1:24,testASEfwd,type="l",xlab="# of predictors",ylab="test vs train ASE",ylim=c(10,50))
index<-which(testASEfwd==min(testASEfwd))
points(index,testASEfwd[index],col="red",pch=10)
rss<-summary(reg.fwd)$rss
lines(1:24,rss/2928,lty=3,col="blue")  #Dividing by 2928 since ASE=RSS/sample size
mtext("ASE - Forward", side = 3, line = -2, outer = TRUE)


par(mfrow=c(1,1))
plot(1:24,testASEbwd,type="l",xlab="# of predictors",ylab="test vs train ASE",ylim=c(10,50))
index<-which(testASEbwd==min(testASEbwd))
points(index,testASEbwd[index],col="red",pch=10)
rss<-summary(reg.bwd)$rss
lines(1:24,rss/2928,lty=3,col="blue")  #Dividing by 2928 since ASE=RSS/sample size
mtext("ASE - Backward", side = 3, line = -2, outer = TRUE)


par(mfrow=c(1,1))
plot(1:24,testASEstp,type="l",xlab="# of predictors",ylab="test vs train ASE",ylim=c(10,50))
index<-which(testASEstp==min(testASEstp))
points(index,testASEstp[index],col="red",pch=10)
rss3<-summary(reg.stp)$rss3
lines(1:24,rss/2928,lty=3,col="blue")  #Dividing by 2928 since ASE=RSS/sample size
mtext("ASE - Stepwise", side = 3, line = -2, outer = TRUE)
```


## Using forward backward, and stepwise to select max number of variables:11
```{r}
# final variable selection for forward variable selection
reg.finalfwd=regsubsets(life_expectancy~.,data=Life,method="forward",nvmax=11)
coef(reg.finalfwd,10)

final.modelfwd<-lm(life_expectancy~ status + year + adult_mortality + percentage_expenditure + measles + bmi + polio + diphtheria + hiv_aids + income_composition_of_resources + schooling, data=Life)

summary(final.modelfwd)

plot(final.modelfwd$fitted.values,Life$life_expectancy,xlab="Predicted",ylab="Life Expectancy",xlim=c(20,110),ylim=c(20,110) )
lines(c(20,110),c(20,110),col="red")
mtext("Forward", side = 3, line = -2, outer = TRUE)




# final variable selection for backward variable selection - the forward and backwards method selected the same model in this case
reg.finalbwd=regsubsets(life_expectancy~. ,data=Life,method="backward",nvmax=11)
coef(reg.finalbwd,10)

final.modelbwd<-lm(life_expectancy~ status + year + adult_mortality + percentage_expenditure + measles + bmi + polio + diphtheria + hiv_aids + income_composition_of_resources + schooling,data=Life)

summary(final.modelbwd)


plot(final.modelbwd$fitted.values,Life$life_expectancy,xlab="Predicted",ylab="Life Expectancy",xlim=c(20,110),ylim=c(20,110))
lines(c(20,110),c(20,110),col="red")
mtext("Backward", side = 3, line = -2, outer = TRUE)




# final variable selection for stepwise variable selection
reg.finalstp=regsubsets(life_expectancy~.,data=Life,method="seqrep",nvmax=11)
coef(reg.finalstp,10)

final.modelstp <- lm(life_expectancy~.-total_expenditure -hepatitis_b -alcohol -infant_deaths, data=Life)

summary(final.modelstp)

plot(final.modelstp$fitted.values,Life$life_expectancy,xlab="Predicted",ylab="Life Expectancy",xlim=c(20,110),ylim=c(20,110) )
lines(c(20,110),c(20,110),col="red")
mtext("Stepwise", side = 3, line = -2, outer = TRUE)


# final variable selection for lasso variable selection
final.modellso <- reg.lso

summary(final.modellso)

plot(final.modellso$fitted.values,Life$life_expectancy,xlab="Predicted",ylab="Life Expectancy",xlim=c(20,110),ylim=c(20,110), na.rm = TRUE )
lines(c(20,110),c(20,110),col="red")
mtext("Lasso", side = 3, line = -2, outer = TRUE)
```


## Comparing the models
```{r}

# Function for Root Mean Squared Error
RMSE <- function(error) { sqrt(mean(error^2)) }

RMSE(final.modelfwd$residuals)
RMSE(final.modelbwd$residuals)
RMSE(final.modelstp$residuals)
RMSE(final.modellso$residuals)


summary(final.modelfwd)$adj.r.squared
summary(final.modelbwd)$adj.r.squared
summary(final.modelstp)$adj.r.squared
summary(final.modellso)$adj.r.squared

AIC(final.modelfwd,final.modelbwd, final.modelstp, final.modellso)
BIC(final.modelfwd,final.modelbwd, final.modelstp, final.modellso)
```
