---
title: "6372 Project 1 - Life Expectancy"
author: "By Kristi Herman, Bo Yun, Eric Romero, Neil Benson"
date: "09/20/2020"
output: word_document 
---
  
  # Introduction
  Introduction paragraphs here

## Importing the necessary libraries
```{r,warning=FALSE,message=FALSE}
library(ggplot2)
library(ggthemes)
library(gridExtra)
library(tidyverse)
library(plyr)
library(dplyr) 
library(naniar) #keep
library(zoo)
# create not in operator
`%notin%` <- Negate(`%in%`)

```


## Importing and Cleaning the Data
```{r,fig.height=6, fig.width=9,echo=FALSE,warning=FALSE,message=FALSE}
########### Loading the data
life_ex_DF <- read.csv("https://raw.githubusercontent.com/kristxh/Applied_Stats_Project1/master/datasets_12603_17232_Life_Expectancy_Data.csv")
# cleaning up column names
colnames(life_ex_DF) <- tolower(colnames(life_ex_DF))
colnames(life_ex_DF) <- gsub("\\.\\.",".",colnames(life_ex_DF))
colnames(life_ex_DF) <- gsub("\\.","_",colnames(life_ex_DF))
names(life_ex_DF)[19] <- "thinness_10_19_years"
# reviewing the structure of the data
str(life_ex_DF)
# converting status to factor column
life_ex_DF$status <- as.factor(life_ex_DF$status)
```


## checking for missing data
```{r}
# checking for missing values in the data
vis_miss(life_ex_DF) + xlab("Data Columns")
```


## fix missing data
```{r, fig.height=9, fig.width=9}
# sorting the columns so that life_expectancy column is first
life_ex_DF <- life_ex_DF %>% select("life_expectancy", everything())
factorcols <- c("status","country","year")
allcols <- colnames(life_ex_DF)
# selecting only numerical columns for correlation
corvars <- allcols[allcols %notin% factorcols]
# impute with median missing values per country
life_ex_median_DF <- as.data.frame(life_ex_DF %>%
                                     group_by(country) %>%
                                     mutate_at(corvars, na.aggregate, FUN=median))
# dropping rows where the country has a single entry/record in the dataset and their records are mostly incomplete
life_ex_median_DF <- life_ex_median_DF[!is.na(life_ex_median_DF$life_expectancy), ]
# imputing the median by country still left a few nulls throughout. We will then impute the rest of the missing values as the mean of stats (developpoing, developed, by year)
life_ex_cleaned_DF <- as.data.frame(life_ex_median_DF %>%
                                      group_by(status,year) %>%
                                      mutate_at(corvars, na.aggregate, FUN=median))
# checking for missing values in the data
vis_miss(life_ex_cleaned_DF) + xlab("Data Columns")
```

# summary of cleaned up data
```{r}
summary(life_ex_cleaned_DF)
```

# correlation matrix with pearson correlation coefficients
```{r}
library(reshape2)
corr <- round(cor(life_ex_cleaned_DF[,corvars]),1)
melted_cormat <- melt(corr)
# Get lower triangle of the correlation matrix
get_lower_tri<-function(cormat){
  cormat[lower.tri(cormat)] <- NA
  return(cormat)
}

# reorder the correlation matrix
reorder_cormat <- function(cormat){
  # Use correlation between variables as distance
  dd <- as.dist((1-cormat)/2)
  hc <- hclust(dd)
  cormat <-cormat[hc$order, hc$order]
}
# reorder the correlation matrix
corr <- reorder_cormat(corr)
# get lower half
lower_half <- get_lower_tri(corr)
# reshape the correlation matrix
melted_cormat <- melt(lower_half, na.rm = TRUE)
# plot it
ggcorrmap <- ggplot(melted_cormat, aes(Var2, Var1, fill = value))+
  geom_tile(color = "white")+
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
                       midpoint = 0, limit = c(-1,1), space = "Lab", 
                       name="Pearson\nCorrelation") +
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 45, vjust = 1, 
                                   size = 8, hjust = 1), axis.title.x = element_blank(), axis.title.y = element_blank())+
  coord_fixed()
ggcorrmap + geom_text(aes(Var2, Var1, label = value), color = "black", size = 2.5)
```


# histogram of life_expectancy
```{r}
theme_set(theme_igray())

le_hist <- geom_histogram(
  mapping = aes(life_expectancy),
  data = life_ex_cleaned_DF,
  bins = 20,
  colour = "gray",
  fill = "#0072B2"
)

ggplot()+ le_hist + xlab("Years") + ggtitle("Life Expectancy")

```

# Overall Life Expectancy
```{r}

le_by_yr <- ddply(life_ex_cleaned_DF, .(year), summarize,  avg_le=mean(life_expectancy))

le_line <- geom_line(
  mapping = aes(x = year, y = avg_le),
  data = le_by_yr,
  size = 1,
  colour = "dodgerblue4"
)

ggplot() + le_line + xlab("Year") + ylab("Life Expectancy") + ggtitle("Average Life Expectancy: 2000 - 2015") 

```

# Developed/Developing Life Expectancy
```{r visual1 fig.width=20, fig.height=10}}
le_by_status <- ddply(life_ex_cleaned_DF, .(year,status), summarize,  avg_le=mean(life_expectancy))
line_colors <- c("dodgerblue4", "#D55E00")

le_line2 <- geom_line(
  mapping = aes(x = year, y = avg_le, group=status, color=status),
  data = le_by_status,
  size = 1
)

ggplot() + le_line2 +  xlab("Year") + ylab("Life Expectancy") + ggtitle("Average Life Expectancy: Developed and Developing Countries") + scale_colour_manual(values=line_colors)

```
# histograms of 

# boxplot by status (developing, developed)
```{r visual1, fig.width=6, fig.height=6}
# Plot boxplot

dev_box <- geom_boxplot(
  mapping = aes(x=status, y=life_expectancy),
  data = life_ex_cleaned_DF,
  colour = "black",
  fill = "#56B4E9",
  groupColors=c('#999999','#E69F00')
)

ggplot()+ dev_box + xlab("Status of Country") + ylab("Life Expectancy") + ggtitle("Life Expectancy: Developed and Developing Countries ")

```


# boxplots by status by year

```{r visual1, fig.width=10, fig.height=6}
# Plot boxplot

yr_box <- stat_boxplot(
  mapping = aes(x=year, y=life_expectancy, group=year),
  data = life_ex_cleaned_DF,
  colour = "black",
  fill = "#56B4E9"
  ) 
  

ggplot()+ yr_box + xlab("Year") + ylab("Life Expectancy") + scale_x_continuous(breaks=c(2000:2015)) + ggtitle("Life Expectancy by Year")  
```
```{r fig.width=10, fig.height=6}
# Scatterplot boxplot

le_scatter <- geom_point(
  mapping = aes(x=year, y=life_expectancy, color=status),
  data = life_ex_cleaned_DF
  ) 
  

ggplot()+ le_scatter + xlab("Year") + ylab("Life Expectancy") + scale_x_continuous(breaks=c(2000:2015)) + ggtitle("Life Expectancy: Developed and Developing Countries")  + scale_colour_manual(values=line_colors)
```